Machine Learning for Mechanistic Models of Metapopulation Dynamics

### 2023-11-01 ms.tex
Initial commit

\documentclass[12pt]{article}

<<setup,echo=F,message=F>>=
# computational intensity, i, also known as run level
i <- 3
library(metapoppkg)
Ncities <- ncol(metapoppkg::incidence) - 1 # first column is date
U <- switch(i,20,Ncities,Ncities)
library(foreach)
library(doParallel)
library(doRNG)
cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset=NA))
if(is.na(cores)) cores <- detectCores()
registerDoParallel(cores)

R0_M6 <- function (params, be = T) 
{
    a = params[ifelse(be, "alpha_be1", "alpha_af1")]
    b = params[ifelse(be, "Beta_be1", "Beta_af1")]
    D = params["D1"]
    mu = params[ifelse(be, "mu_be1", "mu_af1")]
    unname((a + (1 - a) * mu) * b * D)
}

ci_c <- readRDS("ci_c.rds")  # generated by si/si.Rnw
ci_uc <- readRDS("ci_uc.rds") # generated by si/si.Rnw
@

@@

\begin{document}

@@

<<benchmarks,echo=F>>=
benchdir <- paste0("benchmark_",i,"/")
if (!dir.exists(benchdir)) dir.create(benchdir)
stew(file=paste0(benchdir,"iid.rda"),{
  iid_negloglik <- function(theta,x) {
    z <- 0
    for(u in 1:ncol(x)) z <- z - sum(dnbinom(x[,u],
      # variance is mu + mu^2/size
      size=exp(theta["log.size"]),
      mu=exp(theta[u]),log=T))
    return(z)
  }

  if(0) { # currently not used in the ms
    # January 10-23 is observation 1-14
    x1 <- incidence[1:14,2:(U+1)]
    theta1=c(log.mu=log(colMeans(x1)+0.01),log.size=log(2))
    iid_mle1 <- optim(theta1,iid_negloglik,x=x1)

    # January 10- February 3 is observation 1-25
    x2 <- incidence[1:25,2:(U+1)]
    theta2=c(log.mu=log(colMeans(x2)+0.01),log.size=log(2))
    iid_mle2 <- optim(theta2,iid_negloglik,x=x2)
  }

  # January 10- February 8 is observation 1-30
  x3 <- incidence[1:30,2:(U+1)]
  theta3=c(log.mu=log(colMeans(x3)+0.01),log.size=log(2))
  iid_mle3 <- optim(theta3,iid_negloglik,x=x3)

})

stew(file=paste0(benchdir,"ar.rda"),{
  ar_negloglik <- function(theta,x) {
    z <- 0
    for(u in 1:ncol(x)) z <- z - sum(dnbinom(x[,u],
      size=exp(theta["log.size"]),
      mu=exp(theta[u])+exp(theta["log.phi"])*c(0,x[1:(nrow(x)-1),u]),log=T))
    return(z)
  }

  if(0) { # currently not used in the ms
    ar_theta1=c(log.mu=log(0.5*colMeans(x1)+0.01),log.size=log(2),log.phi=log(0.8))
    ar_mle1 <- optim(ar_theta1,ar_negloglik,x=x1)

    ar_theta2=c(log.mu=log(0.5*colMeans(x2)+0.01),log.size=log(2),log.phi=log(0.8))
    ar_mle2 <- optim(ar_theta2,ar_negloglik,x=x2)
  }

  ar_theta3=c(log.mu=log(0.5*colMeans(x3)+0.01),log.size=log(2),log.phi=log(0.8))
  ar_mle3 <- optim(ar_theta3,ar_negloglik,x=x3)

})
@

<<loglik-M5,echo=F>>=
#out3 <- readRDS("j6/out_3/j6.rds")
#out3 <- readRDS("l7/out_3/l7.rds")
out3 <- readRDS("l8/out_3/l8.rds")
mle3 <- unlist(out3[which.max(out3$logLik),-c(1,2)])
lik3 <- unlist(out3[which.max(out3$logLik),c(1,2)])
@

<<loglik-M6,echo=F>>=
#out6 <- readRDS("o5/out_3/o5.rds")
#out6 <- readRDS("s4/out_3/s4.rds")
out6 <- readRDS("s5/out_3/s5.rds")
mle6 <- unlist(out6[which.max(out6$logLik),-c(1,2)])
lik6 <- unlist(out6[which.max(out6$logLik),c(1,2)])
@

<<loglik-M2,echo=F>>=
loglikdir <- paste0("loglik_",i,"/")
if (!dir.exists(loglikdir)) dir.create(loglikdir)
if(0){ # for testing
  library(metapoppkg)
  library(doParallel)
  library(doRNG)
  cores <-  as.numeric(Sys.getenv('SLURM_NTASKS_PER_NODE', unset = NA))
  if(is.na(cores)) cores <- detectCores()
  registerDoParallel(cores)
}
registerDoRNG(234)

li20_rinit <- Csnippet("
    double *S = &S1;
    double *E = &E1;
    double *A = &A1;
    double *I = &I1;
    double *Ca = &Ca1;
    double *Cb = &Cb1;
    double *C = &C1;
    const double *pop = &pop1;
    int u;

    for (u = 0; u < U; u++) {
      E[u] = 0;
      A[u] = 0;
      I[u] = 0;
      Ca[u] = 0;
      Cb[u] = 0;
      C[u] = 0;
      S[u] = nearbyint(pop[u]);
    }

    E[0]=nearbyint(2000*runif(0,1));
    A[0]=nearbyint(2000*runif(0,1));
    for (u = 1; u < U; u++) {
      E[u]=nearbyint(3*mob[14*u][0]*E[0]/(pop[0]));
      A[u]=nearbyint(3*mob[14*u][0]*2000*A[0]/(pop[0]));
    }
")
li20_dunit_measure <- Csnippet("
      double AA,BB;
      double m = C;
      double v = m*m/4 > 4 ? m*m/4 : 4;

      if (cases > m) {
          AA = pnorm(cases-0.5,m,sqrt(v),0,1);
          BB = pnorm(cases+0.5,m,sqrt(v),0,1);
          lik = AA + log(1-exp(BB-AA));
      } else if (cases > 0) {
          AA = pnorm(cases+0.5,m,sqrt(v),1,1);
          BB = pnorm(cases-0.5,m,sqrt(v),1,1);
          lik = AA + log(1-exp(BB-AA));
      } else {
        lik = pnorm(cases+0.5,m,sqrt(v),1,1);
      }
      if(!give_log) lik = exp(lik);
")

## to rebuild the global C definition for mobility, we need to define mob in the recompiled code for rinit
  mob_modify_factor <- 20
  days <- 30

## start: lines 121-136 of metapoppkg/R/li23.R #######################################
  mobi <- metapoppkg::mobility+metapoppkg::v_by_g_day*mob_modify_factor
  popu <- metapoppkg::population[1:days+rep((0:372)*30,each=days),]
  incidence <- metapoppkg::incidence

  # choose the cities with largest number of cases
  index <- order(colSums(incidence)[-1], decreasing = T)[1:U]
  incidence <- incidence[1:days,c(1,index+1)]

  index_mob <- rep(index*14-13,each=14)+rep(0:13,U)
  mobi <- mobi[index_mob,index]
  popu <- popu[rep(index*days-days+1,each=days)+rep(0:(days-1),U),]

  to_C_array <- function(v)paste0("{",paste0(v,collapse=","),"}")
  mobi_rows <- apply(mobi,1,to_C_array)
  mobi_array <- to_C_array(mobi_rows)
  mobi_C <- Csnippet(paste0("const double mob[",U*14,"][",U,"] = ",mobi_array,"; "))
## end: lines 121-136 of metapoppkg/R/li23.R #######################################

model2filters <- bake(file=paste0(loglikdir,"model2.rds"),{
  M2_tmp <- li23(U=U, mob_modify_factor=20,for_ibpf=F,version="li20period3")
  M2 <- spatPomp(M2_tmp,dunit_measure=li20_dunit_measure,rinit=li20_rinit,
    unit_statenames = c('S','E','A','I','Ca','Cb','C'),
    globals = mobi_C
  )
  foreach(k=1:switch(i,2,5,20),.combine=c) %dopar% {
    logLik(bpfilter(M2,block_size=1,Np=switch(i,10,200,2000)))
  }
})
M2_loglik <- logmeanexp(model2filters,se=T)
@

\begin{table}[ht]
\small
\begin{tabular}{c|rr|l}
Model & loglik & df & description
\\
\hline
{\LiMobility} &
  $-\infty$ & 10 &
  SEAIR model using the parameter values and mobility data of \cite{li20}
\\
{\LiParams} &
  \Sexpr{myround(mean(M2_loglik['est']),1)}\stdError{\Sexpr{myround(M2_loglik['se'],1)}}& 10 &
  Adjusted mobility and measurement in {\LiMobility}
\\
{\BenchmarkIID} &
  \Sexpr{myround(-iid_mle3$value,1)}  & \Sexpr{U+1} &
  Independent identically distributed negative binomial
\\
{\BenchmarkAR} &
 \Sexpr{myround(-ar_mle3$value,1)}  & \Sexpr{U+2} &
  Autoregressive negative binomial
\\
{\RevisedModelUnconstrained} &
  \Sexpr{myround(lik3["logLik"],1)}\stdError{\Sexpr{myround(lik3["logLik_se"],1)}}&
  12 &
  Adding overdispersed dynamics to {\LiParams} and refitting
\\
{\RevisedModelConstrained} &
  \Sexpr{myround(lik6["logLik"],1)}\stdError{\Sexpr{myround(lik6["logLik_se"],1)}}&
  10 &
  Latent and infectious durations unchanged by lockdown in {\RevisedModelUnconstrained}.
\\

@@

All parameter estimates are reported in Table~\suppTableResults.
Here, we discuss the estimated basic reproductive number (i.e., the expected number of secondary infections from one index case in a fully susceptible population), denoted by $\Rzero^{\before}$ and $\Rzero^{\after}$ before and after the January 23rd lockdown.
Our estimates for model {\RevisedModelConstrained}, are $\Rzero^{\before}=\Sexpr{myround(ci_c["R0_be", "mle"],2)}$ with confidence interval (CI) (\Sexpr{myround(ci_c["R0_be", "ci1"],2)},\Sexpr{myround(ci_c["R0_be", "ci2"],2)}) and $\Rzero^{\after}=\Sexpr{myround(ci_c["R0_af","mle"],2)}$  with CI (\Sexpr{myround(ci_c["R0_af", "ci1"],2)},\Sexpr{myround(ci_c["R0_af","ci2"],2)}), where the estimates and their associated 95\% CIs obtained by profile likelihood (Supplementary Sec.~\suppSecInference).
This implies that the Chinese government non-pharmaceutical interventions instituted on and around January 23 reduced $\Rzero$ by a factor of $\Sexpr{myround(ci_c["R0_be","mle"]/ci_c["R0_af","mle"],1)}$.
By contrast, the estimates of \cite{li20} are $\widetilde\Rzero^{\before}=2.38$  with CI $(2.03,2.77)$ and $\widetilde\Rzero^{\after}=0.98$  with CI $(0.83,1.16)$, implying reduction by a factor of $\Sexpr{myround(2.38/0.98,1)}$.
For comparison, interventions implemented across a panel of 41 countries (34 European) were estimated to reduce $\Rzero$ by a factor of $\Sexpr{myround(1/(1-0.77),1)}$  with CI $(\Sexpr{myround(1/(1-0.65),1)},\Sexpr{myround(1/(1-0.85),1)})$ \citep{brauner21}.
Our estimate for $\Rzero$ before lockdown is toward the high end of comparable estimates, reviewed by \citep{park20-jcm}.
An alternative metaopopulation analysis of the pre-lockdown China data, with a deterministic transmission model, obtained an $\Rzero$ estimate of 3.11  with CI  (2.39,4.13) \citep{read21}.

>>>>

\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{xcolor}
% maxwidth is the original width if it is less than linewidth
% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%
\let\hlipl\hlkwb

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

\usepackage{alltt}

@@

\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\begin{table}[ht]
\small
\begin{tabular}{c|rr|l}
Model & loglik & df & description
\\
\hline
{\LiMobility} &
  $-\infty$ & 10 &
  SEAIR model using the parameter values and mobility data of \cite{li20}
\\
{\LiParams} &
  -14240.5\stdError{272.4}& 10 &
  Adjusted mobility and measurement in {\LiMobility}
\\
{\BenchmarkIID} &
  -11257.9  & 374 &
  Independent identically distributed negative binomial
\\
{\BenchmarkAR} &
 -10825.3  & 375 &
  Autoregressive negative binomial
\\
{\RevisedModelUnconstrained} &
  -9088.2\stdError{3.4}&
  12 &
  Adding overdispersed dynamics to {\LiParams} and refitting
\\
{\RevisedModelConstrained} &
  -9116.5\stdError{10.1}&
  10 &
  Latent and infectious durations unchanged by lockdown in {\RevisedModelUnconstrained}.
\\

@@

All parameter estimates are reported in Table~\suppTableResults.
Here, we discuss the estimated basic reproductive number (i.e., the expected number of secondary infections from one index case in a fully susceptible population), denoted by $\Rzero^{\before}$ and $\Rzero^{\after}$ before and after the January 23rd lockdown.
Our estimates for model {\RevisedModelConstrained}, are $\Rzero^{\before}=3.51$ with confidence interval (CI) (3.31,3.72) and $\Rzero^{\after}=0.70$  with CI (0.65,0.77), where the estimates and their associated 95\% CIs obtained by profile likelihood (Supplementary Sec.~\suppSecInference).
This implies that the Chinese government non-pharmaceutical interventions instituted on and around January 23 reduced $\Rzero$ by a factor of $5.0$.
By contrast, the estimates of \cite{li20} are $\widetilde\Rzero^{\before}=2.38$  with CI $(2.03,2.77)$ and $\widetilde\Rzero^{\after}=0.98$  with CI $(0.83,1.16)$, implying reduction by a factor of $2.4$.
For comparison, interventions implemented across a panel of 41 countries (34 European) were estimated to reduce $\Rzero$ by a factor of $4.3$  with CI $(2.9,6.7)$ \citep{brauner21}.
Our estimate for $\Rzero$ before lockdown is toward the high end of comparable estimates, reviewed by \citep{park20-jcm}.
An alternative metaopopulation analysis of the pre-lockdown China data, with a deterministic transmission model, obtained an $\Rzero$ estimate of 3.11  with CI  (2.39,4.13) \citep{read21}.
